Nonlinear ${\bs E}$-mode clustering in Lagrangian Space

### 2016-12-18 ms_disp1/ms_disp1.tex
change text according to the referee

\addtolength{\hoffset}{-0.525cm}
\addtolength{\textwidth}{1.05cm}
\title{Lagrangian Space Nonlinear $E$-mode clustering}

@@

Standard BAO reconstruction algorithms apply in Eulerian space.
They usually smooth the nonlinear density field on
linear scale ($\sim$10 Mpc$/h$) and reverse the large 
scale bulk flows by a negative Zel'dovich linear displacement
\citep{2007ApJ...664..675E,2009PhRvD..80l3501N,2009PhRvD..79f3523P}.
In the linear Lagrangian perturbation theory (LPT), the 
negative divergence of the the displacement field
$-\nabla_q\cdot\bs\Psi(\bs q)$ respect to Lagrangian coordinates $\bs q$
gives the linear density field \citep{2010PhDT.........4J},
and in many literatures
\cite{2013MNRAS.428..141N,2014PhRvD..89h3515C,2016JCAP...03..017B}
$\nabla_q\cdot\bs\Psi(\bs q)$ is compared with various standard LPT
and corrected LPT initial conditions. Their motivation was
trying to correct or modify the LPT displacement fields
in order that the final positions of particles are brought
closer to $N$-body results. Because of the absence
of Lagrangian space information in observations, few density
reconstruction algorithms are developed according to the relation
between displacement and linear density. However,
there are techniques to estimate the displacement field from a final 
distribution of matter.
For example, when a homogeneous initial matter distribution 
is assumed, there is a unique solution of curl-less displacement field to relate 
the initial and final distributions without shell-crossing. This solution can be 
solved by a metric transformation equation
\citep{1995ApJS..100..269P,1998ApJS..115...19P}.
In 1-dimensional (1D) case, the exact solution 
simplifies to an ordering of matter elements by Eulerian coordinates.
Zhu {\it et al.} \cite{2016arXiv160907041Z} apply this algorithm to the result
of an 1D simulation \citep{2016JCAP...01..043M} and obtain an estimated
displacement field $\tilde\Psi(q)$, and find
that this new method well recovers the linear
information and reconstructs 1D BAO peak in correlation function.
In 3D and more realistic cases, one needs to
carefully consider effects of curl, shell-crossing, 
complicated baryonic physics and biased tracers (galaxies).

@@

Before these steps, we need to quantify the amount of linear information
that can be recovered\footnote{To avoid ambiguity, we call
$\delta_R\equiv-\nabla_q\cdot\bs\Psi(\bs q)$ recovered linear density field, which
requires $\bs\Psi(\bs q)$ from $N$-body simulations, while
$-\nabla_q\cdot\tilde{\bs \Psi}(\bs q)$ is called the
reconstructed density field, which uses the estimated displacement field.}
 from the full nonlinear displacement field $\bs\Psi(\bs q)$.
In $\bs\Psi(\bs q)$ from $N$-body simulations,
(1) shell-crossing effects are fully considered; (2) $E$-mode displacement
field is used for recovering the linear information (although $B$-mode is
also present); (3) the setup is clean in absence of baryons and realistic
observables. Further reconstructions by $\tilde{\bs \Psi}(\bs q)$
can be compared with this result.
Further more, we study the scale dependent cross-correlation
between $-\nabla_q\cdot\bs\Psi(\bs q)$ and linear density field
and construct optimal Wiener filters to get the optimal filtered
recovered linear density field and recovered linear power spectrum.

@@

\subsection{Simulation}\label{ss.sim}
We use the open source cosmological simulation code {\tt CUBE} %\footnote{github link of cafcube}
\citep{cafcube}.
Cosmological parameters are set as
$\Omega_m=0.27$, $\Omega_\Lambda=0.73$, $h_0=0.68$, $n_s=0.96$ and $\sigma_8=0.83$.
Initial conditions are generated at redshift $z=50$ 
using Zel'dovich approximation, and using a CLASS transfer function
\citep{2011JCAP...07..034B}.
$N_p=512^3$ $N$-body particles are evolved via 
their mutual gravitational interactions to $z=0$, in a periodic box with $L=400$ 
Mpc$/h$ per side. The code is set to use standard a particle-mesh (PM) algorithm 
\cite{1988csup.book.....H} on a two-level mesh grids
(details see \cite{2013MNRAS.436..540H}) and cloud-in-cell
(CIC) is used in particle interpolations in force 
calculation and obtaining the density field $\rho({\bs x})$ in Eulerian coordinates 
${\bs x}$ at late stages. We use density contrast $\delta\equiv\rho/\lb\rho\rb-1$ 
to describe the density fluctuations. The primordial linear density field $
\delta_L$ is given by the initial stage and scaled to $z=0$ by the linear growth 
factor. In the top and middle panels of Fig.1 we show projections of the nonlinear density field
$\delta_N$ given by the simulation and the linear density field $\delta_L$.
$\delta_N$ is obtained by the particle distribution at redshift $z=0$, and
the particles are interpolated by using the cloud-in-cell (CIC) algorithm.
Because $\delta_N$ is highly nonlinear and follows an approximate
log-normal distribution, we plot $\log_{10}(\delta_N+1)$ instead.
The nonlinear evolution of $\delta_N$ makes it very different from $\delta_L$
in appearance.

@@

To quantify the linear information in the recovered density field, we decompose $\delta_L$ in Fourier space as
\begin{equation}\label{eq.decompose}
    \delta_L(k)=r'\delta_R+\delta_n,
\end{equation}
where $r'\delta_R$ is completely correlated with linear density $\delta_L$. Correlating equation (\ref{eq.decompose}) with $\delta_R$ gives
\begin{equation}
    P_{LR}=r'P_{RR}+P_{nR},
\end{equation}
where $P_{ij}\equiv\langle\delta_i\delta_j\rangle$ denotes the cross power 
spectrum. Since $\delta_n$ is uncorrelated with $\delta_R$, $P_{nR}=0$. With the 
definition of cross correlation coefficient $r(\delta_L,\delta_R)\equiv P_{LR}/\sqrt{P_{LL}P_{RR}}$ 
and bias $b^2=P_{RR}/P_{LL}$, we solve $r'=P_{LR}/P_{RR}=rb^{-1}$.
Note that these computations above and below can also be applied on $\delta_N$ for
comparison, by replacing ``$_R$'' with ``$_N$'' in the equations,
while we do not rewrite them explicitly
in the paper for simplicity. From these, we plot the cross 
correlation coefficient $r_{LN}=r(\delta_L,\delta_N)$ and
$r_{LR}=r(\delta_L,\delta_R)$ in Fig.\ref{fig.corr}.
$r_{LN}$ shows no correlation starting from $k\simeq 0.4 h/$Mpc \citep{2004MNRAS.355..129S}.
Clearly, $
\delta_R$ contains much more linear information on smaller scales.

According to equation (\ref{eq.decompose}), the auto power spectrum is decomposed as
\begin{equation}\label{eq.power}
    P_{LL}=r^2b^{-2}P_{RR}+P_{nn},
\end{equation}
and $P_{nn}=(1-r^2)P_{LL}$. Then we construct a Wiener filter to filter out the uncorrelated part in $\delta_R$:
\begin{equation}
    W(k)=\frac{r^2b^{-2}P_{RR}}{r^2b^{-2}P_{RR}+P_{nn}}=r^2.
\end{equation}
The optimal recovered density is given by
\begin{equation}
    \tilde\delta_R=Wb^{-1}\delta_R,
\end{equation}
and the optimal recovered power spectrum is given by
\begin{equation}\label{eq.opt}
    \tilde P=W^2b^{-2}P_{RR}=W^2P_{LL}+W^2b^{-2}P_{NN}.
\end{equation}
Here $W^2$ describes the damping of the linear power spectrum.

@@

The advantage of using displacement field in reconstruction is
its insensitive response from highly nonlinearities.
nonlinear densities $\delta_N$ can be arbitrarily large -- 
one expect virialized regions to be observable, where nonlinear
density is given by the inverse determinant of the Jacobian
$\delta_N=|\bs J|^{-1}\gtrsim 200$.
However, reconstructed densities $\delta_R$ is given by
the trace $\rm{tr}(\bs J)$ and saturates at 3.
Actually, the displacement fields are dominated by early stage linear processes,
which is the Lagrangian-Eulerian coordinate transform,
while late stage shell-crossing, nonlinear and baryonic processes
only fine-tunes the final position $\bs x$
\citep{2014PhRvD..89h3515C}. Compared with estimated displacement fields $\tilde{\bs \Psi}(\bs q)$
by \cite{1995ApJS..100..269P}, which do not have shell crossing,
the additional shell-crossing information in $\bs\Psi(\bs q)$
is uncorrelated with $\delta_L$. This insensitive response from
nonlinearities enables the stability of reconstruction algorithms,
which are expected to give similar results of this paper. 
In contrast, traditional
treatments in reconstruction deals directly on density fields which sensitively
relies on nonlinear processes --  density values can vary by orders
of magnitude due to nonlinear/baryonic physics and many sources of errors.

\section*{Acknowledgements}
We thank Pengjie Zhang for helpful discussions and comments.
We acknowledge funding from NSERC.
HRY acknowledges General Financial Grant No.2015M570884 and Special Financial Grant No.2016T90009 from the China Postdoctoal Science Foundation.
HMZ acknowledges the support of the Chinese MoST 863 program under Grant
No. 2012AA121701, the CAS Science Strategic Priority Research Program
XDB09000000 and the NSFC under Grant No. 11373030.
Research at the Perimeter Institute is supported by the Government of Canada
through Industry Canada and by the Province of Ontario through the Ministry of
Research $\&$ Innovation.

>>>>

\addtolength{\hoffset}{-0.525cm}
\addtolength{\textwidth}{1.05cm}
\title{Nonlinear ${\bs E}$-mode clustering in Lagrangian Space}

@@

Standard BAO reconstruction algorithms apply in Eulerian space.
They usually smooth the nonlinear density field on
linear scale ($\sim$10 Mpc$/h$) and reverse the large 
scale bulk flows by a negative Zel'dovich linear displacement
\citep{2007ApJ...664..675E,2009PhRvD..80l3501N,2009PhRvD..79f3523P}.
In the linear Lagrangian perturbation theory (LPT), the 
negative divergence of the the displacement field
$-\nabla_q\cdot\bs\Psi(\bs q)$ respect to Lagrangian coordinates $\bs q$
gives the linear density field \citep{2010PhDT.........4J},
and in many literatures
\cite{2013MNRAS.428..141N,2014PhRvD..89h3515C,2016JCAP...03..017B}
$\nabla_q\cdot\bs\Psi(\bs q)$ is compared with various standard LPT
and corrected LPT initial conditions. Their motivation was
trying to correct or modify the LPT displacement fields
in order that the final positions of particles are brought
closer to $N$-body results. Because of the absence
of Lagrangian space information in observations, few density
reconstruction algorithms are developed according to the relation
between displacement and linear density. However,
there are techniques to estimate the displacement field from a final 
distribution of matter.
For example, when a homogeneous initial matter distribution 
is assumed, there is a unique solution of curl-less displacement field to relate 
the initial and final distributions without shell-crossing. This solution can be 
solved by a metric transformation equation
\citep{1995ApJS..100..269P,1998ApJS..115...19P}.
In 1-dimensional (1D) case, this solution 
simplifies to an ordering of mass elements by their final, Eulerian coordinates.
Zhu {\it et al.} \cite{2016arXiv160907041Z} apply this algorithm to the result
of an 1D simulation \citep{2016JCAP...01..043M} and obtain an estimated
displacement field $\tilde\Psi(q)$, and find
that this new method well recovers the linear
information and reconstructs 1D BAO peak in correlation function.
In 3D and more realistic cases, one needs to
carefully consider effects of curl, shell-crossing, 
complicated baryonic physics and biased tracers (galaxies).

@@

Before these steps, we need to quantify the amount of linear information
that can be recovered\footnote{To avoid ambiguity, we call
$\delta_R\equiv-\nabla_q\cdot\bs\Psi(\bs q)$ recovered linear density field, which
requires $\bs\Psi(\bs q)$ from $N$-body simulations, while
$-\nabla_q\cdot\tilde{\bs \Psi}(\bs q)$ is called the
reconstructed density field, which uses the estimated displacement field.}
from the full nonlinear displacement field $\bs\Psi(\bs q)$.
This field can be decomposed into a curl-less ``$E$-mode'' component
and a divergence-less ``$B$-mode'' component, $\bs\Psi=\bs\Psi_E+\bs\Psi_B$,
where the $B$-mode is raised by late stage non-Gaussianities and is not
dominant in $\bs\Psi$ \citep{2014PhRvD..89h3515C}.
In $\bs\Psi(\bs q)$ from $N$-body simulations,
(1) $E$-mode displacement field is used for recovering the linear information
(although $B$-mode is also present); (2) shell-crossing effects are fully considered;
(3) the setup is clean in absence of baryons and realistic
observables. Further reconstructions by $\tilde{\bs \Psi}(\bs q)$
can be compared with this result.
Further more, we study the scale dependent cross-correlation
between $-\nabla_q\cdot\bs\Psi(\bs q)$ and linear density field
and construct optimal Wiener filters to get the optimal filtered
recovered linear density field and recovered linear power spectrum.

@@

\subsection{Simulation}\label{ss.sim}
We use the open source cosmological simulation code {\tt CUBE} %\footnote{github link of cafcube}
\citep{cafcube}.
Cosmological parameters are set as
$\Omega_m=0.27$, $\Omega_\Lambda=0.73$, $h_0=0.68$, $n_s=0.96$ and $\sigma_8=0.83$.
Initial conditions are generated at redshift $z=50$ 
using Zel'dovich approximation, and using a CLASS transfer function
\citep{2011JCAP...07..034B}.
$N_p=512^3$ $N$-body particles are evolved via 
their mutual gravitational interactions to $z=0$, in a periodic box with $L=400$ 
Mpc$/h$ per side. The code is set to use standard a particle-mesh (PM) algorithm 
\cite{1988csup.book.....H} on a two-level mesh grids
(details see \cite{2013MNRAS.436..540H}) and cloud-in-cell
(CIC) is used in particle interpolations in force 
calculation and obtaining the density field $\rho({\bs x})$ in Eulerian coordinates 
${\bs x}$ at late stages. We use density contrast $\delta\equiv\rho/\lb\rho\rb-1$ 
to describe the density fluctuations. The primordial linear density field $
\delta_L$ is given by the initial stage and scaled to $z=0$ by the linear growth 
factor. In the top and middle panels of Fig.1 we show projections of the nonlinear density field
$\delta_N$ given by the simulation and the linear density field $\delta_L$.
$\delta_N$ is obtained by the particle distribution at redshift $z=0$, and
the particles are interpolated by using the cloud-in-cell (CIC) algorithm.
Because $\delta_N$ is highly nonlinear and follows an approximate
log-normal distribution, we plot $\log_{10}(\delta_N+1)$ instead, and show
the color scale $\log_{10}(\delta_N+1)\in[-1,1]$ (or $\rho_N/\langle\rho_N\rangle\in[0.1,10]$)
only, for a better visualization.
The nonlinear evolution of $\delta_N$ makes it very different from $\delta_L$
in appearance.

@@

To quantify the linear information in the recovered density field $\delta_R$,
we decompose $\delta_L$ in Fourier space into two uncorrelated parts,
\begin{equation}\label{eq.decompose}
    \delta_L(k)=r'\delta_R+\delta_n,
\end{equation}
where the first term $r'\delta_R$ is completely correlated with $\delta_R$,
meaning the linear information we can extract from $\delta_R$.
The noise term $\delta_n$ is uncorrelated with $\delta_R$, being
the rest of linear information which is not contained in $\delta_R$.
Correlating equation (\ref{eq.decompose}) with $\delta_R$ gives
\begin{equation}
    P_{LR}=r'P_{RR}+P_{nR},
\end{equation}
where $P_{ij}\equiv\langle\delta_i\delta_j\rangle$ denotes the cross power 
spectrum. Since $\delta_n$ is uncorrelated with $\delta_R$, $P_{nR}=0$. With the 
definition of cross correlation coefficient $r(\delta_L,\delta_R)\equiv P_{LR}/\sqrt{P_{LL}P_{RR}}$ 
and bias $b^2=P_{RR}/P_{LL}$, we solve $r'=P_{LR}/P_{RR}=rb^{-1}$.
Note that these computations above and below can also be applied on $\delta_N$ for
comparison, by replacing ``$_R$'' with ``$_N$'' in the equations,
while we do not rewrite them explicitly
in the paper for simplicity. From these, we plot the cross 
correlation coefficient $r_{LN}=r(\delta_L,\delta_N)$ and
$r_{LR}=r(\delta_L,\delta_R)$ in Fig.\ref{fig.corr}.
$r_{LN}$ shows no correlation starting from $k\simeq 0.4 h/$Mpc \citep{2004MNRAS.355..129S}.
Clearly, $
\delta_R$ contains much more linear information on smaller scales.

According to equation (\ref{eq.decompose}), the auto power spectrum of $\delta_L$ is decomposed as
\begin{equation}\label{eq.power}
    P_{LL}=r^2b^{-2}P_{RR}+P_{nn},
\end{equation}
and $P_{nn}=(1-r^2)P_{LL}$. We also explicitly
compute the cross power spectrum between $\delta_R$ and $\delta_n=\delta_L-rb^{-1}\delta_R$,
and found that $rb^{-1}P_{nR}$ is about two orders of magnitude lower than $P_{LL}$,
being consistent with zero. This confirms that
the signal term $rb^{-1}\delta_R$ and the noise term $\delta_n$
is indeed uncorrelated and validates equation (\ref{eq.power}).
According to these two terms, we construct a Wiener filter to filter
out the uncorrelated part in $\delta_R$:
\begin{equation}
    W(k)=\frac{r^2b^{-2}P_{RR}}{r^2b^{-2}P_{RR}+P_{nn}}=r^2.
\end{equation}
The optimal recovered density is given by
\begin{equation}
    \tilde\delta_R=Wb^{-1}\delta_R,
\end{equation}
and the optimal recovered power spectrum is given by
\begin{equation}\label{eq.opt}
    \tilde P=W^2b^{-2}P_{RR}=W^2P_{LL}+W^2b^{-2}P_{NN}.
\end{equation}
Here $W^2$ describes the damping of the linear power spectrum.

@@

The advantage of using displacement field in reconstruction is
its insensitive response from highly nonlinearities.
Nonlinear densities $\delta_N$ can be arbitrarily large -- 
one expect virialized regions to be observable, where nonlinear
density is given by the inverse determinant of the Jacobian
$\delta_N=|\bs J|^{-1}\gtrsim 200$.
However, reconstructed densities $\delta_R$ is given by
the trace $\rm{tr}(\bs J)$ and saturates at 3.
Actually, the displacement fields are dominated by early stage linear processes,
which is the Lagrangian-Eulerian coordinate transform,
while late stage shell-crossing, nonlinear and baryonic processes
only fine-tunes the final position $\bs x$
\citep{2014PhRvD..89h3515C}. Compared with estimated displacement fields $\tilde{\bs \Psi}(\bs q)$
by \cite{1995ApJS..100..269P}, which do not have shell crossing,
the additional shell-crossing information in $\bs\Psi(\bs q)$
is uncorrelated with $\delta_L$. This insensitive response from
nonlinearities enables the stability of reconstruction algorithms,
which are expected to give similar results of this paper. 
In contrast, traditional
treatments in reconstruction deals directly on density fields which sensitively
relies on nonlinear processes --  density values can vary by orders
of magnitude due to nonlinear/baryonic physics and many sources of errors.

\section*{Acknowledgements}
We thank Pengjie Zhang and Kwan Chuen Chan for helpful discussions and comments.
We acknowledge funding from NSERC.
HRY acknowledges General Financial Grant No.2015M570884 and Special Financial Grant No.2016T90009 from the China Postdoctoal Science Foundation.
HMZ acknowledges the support of the Chinese MoST 863 program under Grant
No. 2012AA121701, the CAS Science Strategic Priority Research Program
XDB09000000 and the NSFC under Grant No. 11373030.
Research at the Perimeter Institute is supported by the Government of Canada
through Industry Canada and by the Province of Ontario through the Ministry of
Research $\&$ Innovation.
